# QC analysis of microRNA expression data

```{r setup, include=FALSE}
# source("functions/knit_wiki.r")
# knit_wiki("3c-qc-microrna.Rmd", output.filename = "microRNA data QC analysis")

report.title <- "qc-microrna"
fig.dir <- file.path("figures", report.title)
fig.prefix <- "microrna-"

dir.create(fig.dir, showWarnings = FALSE)

opts_knit$set(upload.fun = function(file, 
  key = "ec071b3a0c69857c0db26a477c1a568c") imgur_upload(file, key))

opts_chunk$set(fig.path = file.path(fig.dir, fig.prefix),
  cache.path = paste(".cache-", report.title, "/", sep = ""),
  echo=FALSE, results = 'hide',
  cache=FALSE, fig.width=8.5, fig.height=5, dpi=150)
```

```{r load-libraries, message=FALSE}
library(affy)
library(ggplot2)
library(plyr)
library(ggaffy)

source("functions/print_xtable.r")
source("functions/save-figures.r")
```

```{r graphing-tweaks}
# Shrink text so labels are legible along x-axis
axis.text.x <- theme_text(size = 5, colour = "grey50", angle = 90, hjust = 1)

wide.opts <- opts(axis.text.x = axis.text.x, legend.position = "top",
  legend.margin = grid::unit(0, "lines"))

# Nice figure label
log.label <- expression(log[2] * " microRNA expression")
```


```{r load-data, cache=FALSE}
load("data/eset-micro.rda")
```


## Distributions of data
Data was log-transformed prior to generating the figures below.

```{r log-transform, echo=TRUE, results='markup'}
exprs(eset.micro) <- log2(exprs(eset.micro))
```

### Quantiles plot
```{r quantiles}
ggaffy_quantiles(eset.micro,
  x = "individual", probs = c(0.01, 0.1, 0.25),
  fill = "batch", shape = "gender", facet.row = "tissue") +
  ylab(log.label) + wide.opts
```

### Boxplot
```{r boxplot}
ggaffy_box(eset.micro, x = "individual", color = "batch", facet.row = "tissue",
  add.outliers = TRUE, outlier.size = .5) +
  ylab(log.label) + wide.opts
```

### Relative expression boxplot
```{r boxplot-rle}
ggaffy_box(eset.micro, x = "individual", method = "rle", color = "batch",
  facet.row = "tissue") + ylab(log.label) + wide.opts
```

### Histogram
```{r histogram, warning=FALSE, fig.width=5, fig.height=8.5}
ggaffy_hist(eset.micro, color = "batch", adjust = 0.5, facet.row = "tissue") +
  wide.opts + xlab(log.label)
```


## Sample clustering
Hierarchical clustering of all samples using euclidean distance produces results very similar to the mRNA expression and metylation data. However, like [the methylation data](Methylation data QC analysis), there is little separation between the cortical regions. 

```{r sample-cluster-heatmap, fig.height=4, fig.width=4.5}
ggaffy_clustmat(eset.micro, color = "tissue", raster = TRUE, 
  dist.method = "euclidean") + 
opts(axis.text.x = theme_blank(), axis.text.y = theme_blank(),
  axis.ticks = theme_blank())
```

## Normalization 
Although the data appears to have been normalized, there are systematic differences in intensity across the different tissues. Should the data be normalized across all tissues prior to looking for differenetial expression between tissues? 

```{r quantiles-across-tissues}
ggaffy_quantiles(eset.micro,
  x = "individual", probs = c(0.01, 0.1, 0.25), facet.col = "tissue") +
  ylab(log.label) + wide.opts + opts(axis.text.x = theme_blank(),
    panel.grid.major = theme_blank(), panel.grid.minor = theme_blank(),
    axis.ticks = theme_blank())
```

## Sample correlations with median reference
The original correlation analysis performed between microRNA and mRNA expression data revealed that many of the strongest correlations were being driven by a small number of outlier samples, particiularly the cerebellar data for individuals 1037, 1076, 827, 4903 and several others. When inspecting the relationship between these samples and a pseudo-reference sample generated from the gene-wise median across all samples within a given tissue, they stand out as outliers.

The following table lists all samples with psuedo-reference correlations that are at least 2 SDs below the tissue-specific mean. 

```{r median-ref-correlations, results='asis'}
tissues <- levels(eset.micro$tissue)

ref.cors <- list()
for(t in tissues) {
  micro.sub <- eset.micro[fData(eset.micro)$chr %in% 1:22, eset.micro$tissue == t]
  exp.mat <- exprs(micro.sub)
  med.ref <- rowMedians(exp.mat)
  cors.sub <- apply(exp.mat, 2, cor, y = med.ref, use = "pairwise.complete")
  cors.sub <- data.frame(tissue = t, sample = sampleNames(micro.sub),
    individual = micro.sub$individual, r = cors.sub)
  ref.cors[[t]] <- cors.sub
}

ref.cors <- data.frame(do.call("rbind", ref.cors), row.names = NULL)

# Calculate outlier threshold
thresh <- mean(ref.cors$r) - (sd(ref.cors$r) * 2)

# Table of outlier samples
ref.cors$outlier <- ifelse(ref.cors$r < thresh, TRUE, FALSE)
print_xtable(subset(ref.cors, outlier, select = -outlier), col.names = TRUE)

# Export correlations
write.csv(ref.cors, paste(file.path(fig.dir, fig.prefix), 
  "median-ref-correlations.csv", sep = ""), row.names = FALSE)
```


### Cerebellum median reference correlations
```{r median-ref-correlations-crblm, fig.height = 8.5}
micro.sub <- eset.micro[, eset.micro$tissue == "CRBLM"]
sampleNames(micro.sub) <- micro.sub$individual
ggaffy_scatgrid(micro.sub)
```

### Pons median reference correlations
```{r median-ref-correlations-pons, fig.height = 8.5}
micro.sub <- eset.micro[, eset.micro$tissue == "PONS"]
sampleNames(micro.sub) <- micro.sub$individual
ggaffy_scatgrid(micro.sub)
```

### Frontal cortex median reference correlations
```{r median-ref-correlations-fctx, fig.height = 8.5}
micro.sub <- eset.micro[, eset.micro$tissue == "FCTX"]
sampleNames(micro.sub) <- micro.sub$individual
ggaffy_scatgrid(micro.sub)
```

### Temporal cortex
```{r median-ref-correlations-tctx, fig.height = 8.5}
micro.sub <- eset.micro[, eset.micro$tissue == "TCTX"]
sampleNames(micro.sub) <- micro.sub$individual
ggaffy_scatgrid(micro.sub)
```

## Summary

* Remove `r I(sum(ref.cors$outlier))` samples with psuedo-reference correlations less than 2 SDs below the mean.