# Correlations between microRNAs and mRNA predicted targets

```{r knitr-setup, include=FALSE}
report.title <- "crossdata-correlations-micro-exp"
fig.path <- file.path("figures", report.title)
fig.prefix <- "cors-micro-exp-"

dir.create(fig.path, showWarnings = FALSE)

opts_chunk$set(dev='jpeg', fig.path = file.path(fig.path, fig.prefix),
  echo=FALSE, results = 'hide',
  cache=FALSE, fig.width=3, fig.height=3, dpi=150)
```

```{r load-libraries, cache=FALSE, message=FALSE}
library(Biobase)
library(grid)
library(plyr)
library(ggplot2)
 
source("functions/knit_template.r")
source("functions/print_xtable.r")
source("functions/save-figures.r")
```

```{r load-data, cache=FALSE}
load("data/crossdata-correlations-micro-exp/eset-exp-adjusted.rda")
load("data/crossdata-correlations-micro-exp/eset-micro-adjusted.rda")
```

Explore and visualize correlation results generated between the microRNA and mRNA expression data based on various target prediction resources.

* data adjustment: [`crossdata-correlations-micro-exp-adjust-data.r`][1]
* correlation analysis: [`crossdata-correlations-micro-exp-analysis.r`][2]


## Methods
The [analysis][2] was performed independently for each tissue. Subsets of the expression and methylation eSet objects were created to include only samples from individuals present in both datasets. mRNA and microRNA expression levels were [adjusted][1] for age and sample tissue bank by fitting a linear model with these variables and using the residuals in the correlation analysis.

Predicted microRNA/mRNA interactions were obtained using the various databases provided by the `RmiR.hsa` package. Correlations were calculated using the standard `cor` function and with use set to `"pairwise.complete"`. P-values were obtained using the `WGCNA` package's `corPvalueStudent` function and multiple testing corrections were performed using the Benjamini & Hochberg method for generating q-values.


## Results
Correlation results are provided independently for each microRNA/mRNA target prediction database. 

```{r generate-report, include=FALSE}
# Identify files correlation results from each RmiR.hsa database
mirdbs <- list.files("results", pattern = "crossdata-micro-exp-correlations")
mirdbs <- gsub(".*-(\\w+)\\.rda$", "\\1", mirdbs)

report <- knit_template("templates/crossdata-correlations-micro-exp-report.tpl",
  labels = mirdbs, variable = "mirdb")
```

```{r print-report, results='asis', message=FALSE}
cat(report, "\n")
```

## Summary 

### Outlier exclusion
The original analysis was run using mirbase predictions and included all samples for which microRNA and mRNA expression data was available. The original results can be obtained from commit [468768aa9d](https://github.com/aaronwolen/gibbs-brain-project/commit/468768aa9d). The analysis was then re-run removing samples that showed extreme deviations from a median-generated reference sample (see results from the QC analyses for more information). Without these outliers, the [updated results](https://github.com/aaronwolen/gibbs-brain-project/commit/e404d0a) contained ~2,000 fewer correlations in the cerebellum with a q-value < 0.05. However, at the same level of significance, there were ~2,000 more in pons, ~3,000 more in temporal cortex and ~4,000 more in the frontal cortex.

[1]: https://github.com/aaronwolen/gibbs-brain-project/blob/master/crossdata-correlations-micro-exp-adjust-data.r "crossdata-correlations-micro-exp-adjust-data.r"
[2]: https://github.com/aaronwolen/gibbs-brain-project/blob/master/crossdata-correlations-micro-exp-analysis.r "crossdata-correlations-micro-exp-analysis.r"

### Directionality
In all tissues the number of positive correlations exceeded that of negative correlations.

### Correlation frequency
As shown in the boxplots, there is a subset of microRNAs that are correlated with very large lists of genes.