# Correlation of CpG methylation levels across tissues 

```{r setup, include=FALSE}
# source("functions/knit_wiki.r")
# knit_wiki("crosstissue-correlations-meth-report.Rmd", output.filename = "Cross-tissue methylation correlations")

report.title <- "crosstissue-correlations-meth"
fig.dir <- file.path("figures", report.title)
fig.prefix <- "cors-meth-"

dir.create(fig.dir, showWarnings = FALSE)

opts_knit$set(upload.fun = function(file, 
  key = "ec071b3a0c69857c0db26a477c1a568c") imgur_upload(file, key))

opts_chunk$set(fig.path = file.path(fig.dir, fig.prefix),
  cache.path = paste(".cache-", report.title, "/", sep = ""),
  echo=FALSE, results = 'hide',
  cache=FALSE, fig.width=5.5, fig.height=4, dpi=150)
```

```{r load-libraries, message=FALSE}
# library(Biobase)

library(grid)
library(plyr)
library(ggplot2)
library(reshape2)
 
source("functions/print_xtable.r")
source("functions/save-figures.r")
devtools::source_gist("3298388")
source("functions/heat_matrix.r")
```

```{r load-data, cache=FALSE, message=FALSE}
load("data/crosstissue-correlations-meth/eset-meth-adjusted.rda")

cors.df <- read.csv("results/crosstissue-correlations-meth.csv",
  stringsAsFactors = FALSE)
cors.df$tissue.pair <- factor(cors.df$tissue.pair)

# Annotate probes
f.cols <- c("symbol", "chr", "position", "strand", "cpg.island", 
  "distance.to.tss", "description")

cors.df <- data.frame(cors.df, 
  fData(eset.meth)[cors.df$id, f.cols],
  row.names = NULL)
cors.df <- transform(cors.df, distance.to.tss = as.numeric(distance.to.tss))
```

Explore and visualize cross-tissue correlations results. 

* script: [`crosstissue-correlations-meth-analysis.r`](https://github.com/aaronwolen/gibbs-brain-project/blob/master/crosstissue-correlations-meth-analysis.r)
* results: [`crosstissue-correlations-meth.csv`](https://github.com/aaronwolen/gibbs-brain-project/blob/master/results/crosstissue-correlations-meth.csv)


## Methods
This analysis was restricted to only autosomal probes. Low-variance probes were filtered out by removing any probes with an experiment-wide standard deviation < 0.05.

Methylation levels were adjusted for age and sample tissue bank by fitting a linear model with these variables and using the residuals in the correlation analysis. 

Correlations were calculated using the standard `cor` function and with use set to `"pairwise.complete"`. P-values were obtained using the `WGCNA` package's `corPvalueStudent` function and multiple testing corrections were performed using the Benjamini & Hochberg method for generating q-values within each tissue pair.

## Results

### Table of significant cross-tissue correlations
```{r results-table, results='asis'}
results.table <- ddply(cors.df, .(tissue.pair), summarise, 
  'p-value < 0.05' = sum(pvalue < 0.05),
  'q-value < 0.05' = sum(qvalue < 0.05))

print_xtable(results.table)
```

### Correlation and p-value histograms
Histograms of the Pearson correlation values (left) and corresponding p-values (right).

```{r histograms, message=FALSE, fig.height=5}
p.hist <- ggplot(cors.df) + 
  aes(x = pvalue) + geom_histogram() +
  facet_grid(tissue.pair~.)

r.hist <- ggplot(cors.df) + 
  aes(x = r) + geom_histogram() +
  geom_vline(xintercept = 0, color = "white") +
  facet_grid(tissue.pair~.)

ggaffy:::ggaffy_layout(nrow = 1, ncol = 2, plot.list = list(r.hist, p.hist))
```


### Pairwise comparisons of tissue-pair correlations

```{r r-tissue-pairs, fig.height=5}
r.mat <- dcast(data = cors.df, 
  id + chr + cpg.island + distance.to.tss ~ tissue.pair, value.var = "r")


ggpairs(r.mat, facet.vars = levels(cors.df$tissue.pair), 
  facet.scale = "fixed", density = TRUE) + coord_fixed() + 
  xlab("Cross-tissue correleation") + 
  ylab("Cross-tissue correlation")
```


#### Correlation of cross-tissue methylation correlations
The following heatmap depicts the cross-tissue correlations.

```{r r-tissue-pairs-correlations, fig.width=3.5, fig.height=2.5}
r.cors <- cor(r.mat[, levels(cors.df$tissue.pair)])
dimnames(r.cors) <- lapply(dimnames(r.cors), function(x) sub("-", "\n", x))

heat_matrix(r.cors, symmetric = TRUE, legend.title = "cor", font.size = 3)
```

### Cross-tissue pairs plots for significant correlations
Pairs plots for the 50 most significant cross-tissue correlations are provided [here](https://github.com/aaronwolen/gibbs-brain-project/blob/master/figures/crosstissue-correlations-meth/cors-meth-top-probes.pdf).

```{r sig-cors-pairs-plot, include=FALSE}

# Produce visualizations for individual probes with largest correlations
sig.cors <- subset(cors.df, qvalue < .05)
sig.cors <- arrange(sig.cors, qvalue)

# Top 50 unique probes
probes <- unique(sig.cors$id)[1:50]

probe.plots <- list()
for(p in probes) {

  p.df <- as.data.frame.table(exprs(eset.meth)[p,], stringsAsFactors = F)
  p.df <- cbind(p.df, pData(eset.meth)[p.df$Var1, c("tissue", "individual")])
  p.df <- dcast(p.df, individual ~ tissue, value.var = "Freq")
  p.df <- join(p.df, pData(eset.meth), by = "individual", match = "first")
  
  p.title <- with(subset(sig.cors, id == p)[1,],
    paste(symbol, " (", tissue.pair, " cor = ", round(r, 2), ")", sep = ""))
  
  probe.plots[[p]] <- ggpairs(p.df, mapping = aes(color = gender),
    facet.vars = levels(eset.meth$tissue), facet.scale = "fixed") +
  stat_smooth(method = "lm") +
  xlab(paste("CpG methylation (", p, ")", sep = "")) +
  ylab(paste("CpG methylation (", p, ")", sep = "")) +
  opts(title = p.title) + coord_fixed()
}

pdf(file.path(fig.dir, paste(fig.prefix, "top-probes.pdf", sep = "")),
    width = 7, height = 6)
temp <- lapply(probe.plots, print)
dev.off()
```
