# Individual variation in methylation levels within tissues

```{r setup, include=FALSE}
# source("functions/knit_wiki.r")
# knit_wiki("individual-variation-meth.Rmd", output.filename = "Individual variation in methylation levels")

report.title <- "individual-variation-meth"
fig.dir <- file.path("figures", report.title)
fig.prefix <- "meth-var-"

dir.create(fig.dir, showWarnings = FALSE)

# opts_knit$set(upload.fun = function(file, 
#   key = "ec071b3a0c69857c0db26a477c1a568c") imgur_upload(file, key))

opts_chunk$set(fig.path = file.path(fig.dir, fig.prefix),
  cache.path = paste(".cache-", report.title, "/", sep = ""),
  echo=FALSE, results = 'hide',
  cache=FALSE, autodep=TRUE, fig.width=5, fig.height=4, dpi=150)
```

```{r load-libraries, cache=FALSE, message=FALSE}
library(plyr)
library(reshape2)

library(ggaffy)
library(ggplot2)
library(grid)

source("functions/save-figures.r")
source("functions/plot_matrix.r")
source("functions/heat_matrix.r")
```

```{r load-data, cache=FALSE, message=FALSE}
source("4b-load-meth-data.r")

fData(eset.meth) <- transform(fData(eset.meth), 
  distance.to.tss = as.numeric(distance.to.tss))
```

Individual variation in CpG methylation was observed by calculating each probe's SD across all samples within a tissue.

## Probe variation across individuals
As the shown in the figures below, the distrubtions of probe SDs are comparable across tissues, although variation is highest in the cerebellum. As expected, variation is much lower for probes targetting loci within CpG islands. 

```{r probe-sd-distributions, warning=FALSE}

# Calculate summary statistics for each probe, within tissue
probe_summary <- function(x, tissue) {
  out <- data.frame(id = rownames(x), tissue = tissue,
    mean = rowMeans(x, na.rm = TRUE), 
    median = rowMedians(x, na.rm = TRUE),
    sd = apply(x, 1, sd))
  return(out)
}

probe.df <- list()
for(t in levels(eset.meth$tissue)) {
  probe.df[[t]] <- probe_summary(exprs(eset.meth[, eset.meth$tissue == t]), t)
}

probe.df <- data.frame(do.call("rbind", probe.df), row.names = NULL)

# Annotate probe summaries
f.cols <- c("id", "chr", "color.channel", "symbol", 
  "range.start", "distance.to.tss", "cpg.island")

probe.df <- merge(fData(eset.meth)[, f.cols], probe.df, by = "id", all.x = F)
probe.df$autosome <- ifelse(probe.df$chr %in% 1:22, TRUE, FALSE)

# Density histogram of probe variation across tissues
sd.hist <- ggplot(probe.df) + 
  aes(x = sd, color = tissue) +
  geom_density(alpha = .5) +
  xlab("Probe SD") + opts(legend.position = "top")

sd.bp <- ggplot(probe.df) + 
  aes(x = tissue, y = sd, fill = cpg.island) +
  geom_boxplot(outlier.size = 1) +
  ylab("Probe SD") +
  opts(legend.position = "top")

ggaffy:::ggaffy_layout(nrow = 1, ncol = 2, plot.list = list(sd.hist, sd.bp))
```

## Comparison of methylation variation across tissues

The following plot includes only autosomal probes. 
```{r probe-sd-tissue-pairs, fig.height=5, warning=FALSE}
probe.mat <- dcast(probe.df, id ~ tissue, value.var = "sd")
probe.mat <- merge(fData(eset.meth)[, f.cols], probe.mat, by = "id", all.x = F)

probe.mat <- arrange(probe.mat, desc(cpg.island))

plot_matrix(subset(probe.mat, chr %in% 1:22), density = TRUE, bins = 75,
  facet.vars = levels(eset.meth$tissue)) +
xlab("Probe SD") + ylab("Probe SD") +
opts(legend.position = "top",
  axis.text.x = theme_text(angle = 90, hjust = 1, colour = "grey50", 
    size = base_size * 0.8))

```

Pairwise correlations among tissues:
```{r probe-sd-tissue-correlations, fig.width = 4, fig.height = 3}
heat_matrix(cor(probe.mat[,levels(eset.meth$tissue)], use = "pairwise"),
            symmetric = TRUE, font.size = 3)
```


## Methylation variation as a function of TSS distance

```{r tss-distance, warning=FALSE}

# Bin distance.to.tss
tss.range <- range(pretty(fData(eset.meth)$distance.to.tss))
tss.breaks <- seq(tss.range[1], tss.range[2], 25)
tss.axis <- seq(min(tss.breaks), max(tss.breaks), 100)

# Add bins to fdata
fData(eset.meth)$tss.bin <- cut(fData(eset.meth)$distance.to.tss, 
  breaks = tss.breaks, labels = F)

# Convert to factor
fData(eset.meth)$tss.bin <- factor(fData(eset.meth)$tss.bin, 
  levels = sort(unique(fData(eset.meth)$tss.bin)),
  labels = tail(tss.breaks, -1))

# Add to probe.df
probe.df$tss.bin <- fData(eset.meth)$tss.bin[match(probe.df$id, 
  fData(eset.meth)$id)]
  
ggplot(subset(probe.df, !is.na(tss.bin))) + 
  aes(x = tss.bin, y = sd, color = tissue, group = tissue) +
  stat_summary(fun.y = mean, geom = "line") +
  stat_summary(fun.data = mean_cl_boot, geom = "pointrange") +
  scale_x_discrete(breaks = tss.axis, labels = tss.axis) +
  facet_grid(cpg.island ~., labeller = label_both) +
  xlab("Distance from TSS (nt)") + ylab("CpG methylation") +
  opts(legend.position = "top",
    axis.text.x = theme_text(angle = 90, colour = "grey50", hjust = 1, 
                             size = base_size * 0.8))
```



```{r individual-probe-plots, eval=FALSE}
# Pairs plot of methylation probes across tissues
plot.df <- get_eset_data(eset.meth, probesets = "cg14159672", shape = "long")

plot.df <- dcast(plot.df,
  tissuebank + batch + age + gender + individual ~ tissue, 
  value.var = "expression")
 
plot_matrix(plot.df, facet.vars = levels(eset.meth$tissue), 
  mapping = aes(color = age, shape = gender), size = 2, alpha = 0.5)
```  


##  Variation in GABA receptor methylation

```{r gaba-data}
# Identify probes targeting gaba receptor genes
gaba.probes <- featureNames(eset.meth)[grep("GABR",fData(eset.meth)$symbol)]

# Construct gaba receptor methylation data.frame 
gaba.df <- get_eset_data(eset.meth, probesets = gaba.probes, shape = "long")

gaba.df$symbol <- factor(gaba.df$symbol, 
  levels = sort(as.character(unique(gaba.df$symbol))))

# Nice labels for the different subunits
gaba.df$subunit <- factor(gsub("(GABR\\w{1}).*", "\\1", gaba.df$symbol))
```

### GABA receptor boxplot
```{r gabar-boxplot, fig.width=8.5, warning=FALSE}
ggplot(gaba.df) +
  aes(x = id, y = expression, fill = subunit) +
  geom_boxplot(outlier.size = 1) + 
  ylab("Methylation") +
  facet_grid(tissue ~ subunit, scales = "free_x", space = "free_x") + 
  opts(legend.position = "none",
    axis.text.x = theme_text(angle = 90, colour = "grey50", hjust = 1, 
                                size = base_size * 0.8))
```

### Clustering of GABA receptors
The following heatmaps were constructed by performing average hierarchical clustering of GABA receptors using Pearson's correlate distance. GABA receptor genes are colored by subtype.

```{r gabar-heatmap, fig.width=8.5, fig.height=8}
# Generate clustered heatmap gaba receptors for each tissue
gaba.hms <- list()

for(t in levels(eset.meth$tissue)) {
  eset.sub <- eset.meth[, eset.meth$tissue == t]
  
  # Add GABR classes
  fData(eset.sub)$GABA <- gsub("(GABR\\w{1}).*", "\\1", fData(eset.sub)$symbol)
  fData(eset.sub)$GABA <- factor(fData(eset.sub)$GABA)

  # Only include samples with no missing values
  samples <- apply(exprs(eset.sub)[gaba.probes,], 2, function(x) sum(is.na(x)))
  samples <- sampleNames(eset.sub)[which(samples == 0)]

  gaba.hm <- ggaffy_clustmat(eset.sub, what = "rows",
    samples = samples, probesets = gaba.probes, color = "GABA")

    # Swap probesets for symbols
    gaba.hm <- gaba.hm + 
      scale_x_discrete(breaks = gaba.df$probeset, labels = gaba.df$symbol) +
      scale_y_discrete(breaks = gaba.df$probeset, labels = gaba.df$symbol) +
      scale_fill_gradient2(midpoint = 1) +
      opts(title = t, legend.position = "none")
    gaba.hms[[t]] <- gaba.hm
}

ggaffy:::ggaffy_layout(2, 2, gaba.hms)
```
