# Individual variation in methylation levels within tissues

```{r setup, include=FALSE}
# source("functions/knit_wiki.r")
# knit_wiki("individual-variation-meth.Rmd", output.filename = "Individual variation in methylation levels")

report.title <- "individual-variation-meth"
fig.dir <- file.path("figures", report.title)
fig.prefix <- "meth-var-"

dir.create(fig.dir, showWarnings = FALSE)

opts_knit$set(upload.fun = function(file, 
  key = "ec071b3a0c69857c0db26a477c1a568c") imgur_upload(file, key))

opts_chunk$set(fig.path = file.path(fig.dir, fig.prefix),
  cache.path = paste(".cache-", report.title, "/", sep = ""),
  echo=FALSE, results = 'hide',
  cache=TRUE, autodep=TRUE, fig.width=5, fig.height=3, dpi=150)
```

```{r load-libraries, cache=FALSE, message=FALSE}
library(data.table)
library(plyr)
library(reshape2)

library(ggaffy)
library(ggplot2)
library(grid)

# ggpairs function
devtools::source_gist("3298388")

source("functions/save-figures.r")
source("functions/heat_matrix.r")
source("functions/print_xtable.r")
```

```{r load-data, cache=FALSE, message=FALSE}
source("4b-load-meth-data.r")

fData(eset.meth) <- transform(fData(eset.meth), 
  distance.to.tss = as.numeric(distance.to.tss))
```

Individual variation in CpG methylation was examined by calculating each probe's SD across all samples within a tissue.

## Probe variation across individuals
As the shown in the figures below, the distrubtions of probe SDs are comparable across tissues, although variation is highest in the cerebellum. As expected, variation is much lower for probes targetting loci within CpG islands. 

```{r setup-data-tables, warning=FALSE}
# phenotype data.table for sample annotation
p.cols <- c("geo", "tissue", "gender", "age")
pdata <- data.table(pData(eset.meth)[, p.cols])
setkey(pdata, geo)

# data.table for probe annotation
f.cols <- c("id", "chr", "color.channel", "symbol", 
  "range.start", "distance.to.tss", "cpg.island")
fdata <- data.table(fData(eset.meth)[, f.cols])
setkey(fdata, id)

# Create long-format data.table of methylation data
meth.df <- as.data.frame.table(exprs(eset.meth), stringsAsFactors = FALSE)
names(meth.df) <- c("id", "geo", "expression")

meth.df <- data.table(meth.df)
setkey(meth.df, geo, id)

# Remove NA values
meth.df <- na.omit(meth.df, expression)

# Annotate samples
meth.df <- merge(meth.df, pdata)
setkey(meth.df, tissue, gender, id)
```

```{r probe-sd-distributions}
# Calculate summary statistics for each probe, within tissue
probe.df <- meth.df[, 
  list(mean = mean(expression), 
       median = median(expression),
       sd = sd(expression),
       n = .N),
  by = "tissue,id"]

# Annotate probe summaries
probe.df <- merge(probe.df, fdata, by = "id")

# Flag autosomal probes
probe.df$autosome <- ifelse(probe.df$chr %in% 1:22, TRUE, FALSE)

# Density histogram of probe variation across tissues
sd.hist <- ggplot(probe.df) + 
  aes(x = sd, color = tissue) +
  geom_density(alpha = .5) +
  xlab("Probe SD") + opts(legend.position = "top")

sd.bp <- ggplot(probe.df) + 
  aes(x = tissue, y = sd, fill = cpg.island) +
  geom_boxplot(outlier.size = 1) +
  ylab("Probe SD") +
  opts(legend.position = "top")

ggaffy:::ggaffy_layout(nrow = 1, ncol = 2, plot.list = list(sd.hist, sd.bp))
```

## Comparison of methylation variation across tissues

The following plot includes only autosomal probes. 
```{r probe-sd-tissue-pairs, fig.height=5, warning=FALSE}
sd.mat <- dcast(probe.df, id ~ tissue, value.var = "sd")
sd.mat <- merge(fdata, sd.mat, by = "id", all.x = F)

ggpairs(data.frame(subset(sd.mat, chr %in% 1:22)), 
  density = TRUE, bins = 50,
  facet.vars = levels(eset.meth$tissue)) +
xlab("SD (autosomal probes)") + ylab("SD (autosomal probes)") +
opts(legend.position = "top",
  axis.text.x = theme_text(angle = 90, hjust = 1, colour = "grey50", 
    size = base_size * 0.8))
```


Probe SD cross-tissue correlations:
```{r probe-sd-tissue-correlations, fig.width=4}
cor.mat <- cor(data.frame(sd.mat)[,levels(eset.meth$tissue)])
heat_matrix(cor.mat, symmetric = TRUE, font.size = 3)
```


## Methylation variation as a function of TSS distance

The following plot shows how variability in CpG methylation changes as a function of distance to the nearest transcription start site (TSS). TSS distances were split into 100 nt bins. Each point represents the mean of SDs for all probes grouped into each of the bins plotted along the x-axis. Error bars provide the 95% confidence intervals for each bin, which were bootstrapped 1,000 times. 

```{r tss-distance, warning=FALSE, fig.width=6, fig.height=4}

# Bin distance.to.tss
tss.range <- range(pretty(fdata$distance.to.tss))
tss.breaks <- seq(tss.range[1], tss.range[2], 25)
tss.axis <- seq(min(tss.breaks), max(tss.breaks), 100)

# Add bins to fdata
fdata$tss.bin <- cut(fdata$distance.to.tss, breaks = tss.breaks, labels = F)

# Convert to factor
fdata$tss.bin <- factor(fdata$tss.bin, 
  levels = sort(unique(fdata$tss.bin)),
  labels = tail(tss.breaks, -1))

# Add to probe.df
probe.df$tss.bin <- fdata$tss.bin[match(probe.df$id, fdata$id)]

ggplot(subset(probe.df, !is.na(tss.bin))) + 
  aes(x = tss.bin, y = sd, color = tissue, group = tissue) +
  stat_summary(fun.y = mean, geom = "line") +
  stat_summary(fun.data = mean_cl_boot, geom = "pointrange") +
  scale_x_discrete(breaks = tss.axis, labels = tss.axis) +
  facet_grid(cpg.island ~., labeller = label_both) +
  xlab("Distance to TSS (nt)") + ylab("Mean probe SD") +
  opts(legend.position = "top",
    axis.text.x = theme_text(angle = 90, colour = "grey50", hjust = 1, 
                             size = base_size * 0.8))
```



```{r individual-probe-plots, eval=FALSE}
# Pairs plot of methylation probes across tissues
plot.df <- get_eset_data(eset.meth, probesets = "cg14159672", shape = "long")

plot.df <- dcast(plot.df,
  tissuebank + batch + age + gender + individual ~ tissue, 
  value.var = "expression")
 
ggpairs(plot.df, facet.vars = levels(eset.meth$tissue), 
  mapping = aes(color = age, shape = gender), size = 2, alpha = 0.5)
```  


## Methylation variation as a function of age

```{r create-age-groups, results='asis', message=FALSE}
# Add age groups
age.breaks <- quantile(pdata$age)
pdata$age.group <- cut(pdata$age, breaks = age.breaks, include.lowest = TRUE)

print_xtable(
  with(subset(pdata, gender == "female"), table(tissue, age.group)),
  row.names = TRUE, caption = "Number of females in each age group.", 
  caption.placement = "top")

print_xtable(
  with(subset(pdata, gender == "male"), table(tissue, age.group)),
  row.names = TRUE, caption = "Number of males in each age group.", 
  caption.placement = "top")

# Add age.group to meth.df
meth.df <- merge(meth.df, pdata[, geo,age.group], by = "geo")

# Calculate summary statistics for each level of the phenotype variables
probe.df <- meth.df[, 
  list(mean = mean(expression), 
       median = median(expression),
       sd = sd(expression),
       n = .N),
  by = "tissue,gender,age.group,id"]
setkey(probe.df, id)
```

```{r sd-age-boxplot, fig.width=7.5, fig.height=2.5}
ggplot(probe.df) + 
  aes(x = age.group, y = sd) + 
  geom_boxplot(outlier.size = 1) + 
  xlab("Age group") + ylab("Probe SD") +
  facet_grid(. ~ tissue)
```

## Methylation variation as a function of age and gender

```{r sd-age-gender-boxplot, fig.width=7.5}
ggplot(probe.df) + 
  aes(x = age.group, y = sd, fill = gender) + 
  geom_boxplot(outlier.size = 1) + 
  xlab("Age group") + ylab("Probe SD") +
  facet_grid(. ~ tissue) +
  opts(legend.position = "top")
```

##  Variation in GABA receptor methylation

```{r gaba-data}
# Identify probes targeting gaba receptor genes
gaba.probes <- subset(fdata, grepl("GABR", symbol))$id
gaba.df <- meth.df[id %in% gaba.probes]

# Annotate gaba probes
gaba.df <- merge(gaba.df, fdata, by = "id")

# Convert gaba symbols to ordered factor
gaba.df$symbol <- factor(gaba.df$symbol, 
  levels = sort(as.character(unique(gaba.df$symbol))))

# Nice labels for the different subunits
gaba.df$subunit <- factor(gsub("(GABR\\w{1}).*", "\\1", gaba.df$symbol))
```

### GABA receptor boxplot
```{r gabar-boxplot, fig.width=7.5, fig.height=5, warning=FALSE}
ggplot(gaba.df) +
  aes(x = id, y = expression, fill = subunit) +
  geom_boxplot(outlier.size = 1) + 
  ylab("Methylation") +
  facet_grid(tissue ~ subunit, scales = "free_x", space = "free_x") + 
  opts(legend.position = "none",
    axis.text.x = theme_text(angle = 90, colour = "grey50", hjust = 1, 
                                size = base_size * 0.8))
```

### Clustering of GABA receptors
The following heatmaps were constructed by performing average hierarchical clustering of GABA receptors using Pearson's correlate distance. GABA receptor genes are colored by subtype.

```{r gabar-heatmap, fig.width=7.5, fig.height=7}
# Generate clustered heatmap gaba receptors for each tissue
gaba.hms <- list()

for(t in levels(eset.meth$tissue)) {
  eset.sub <- eset.meth[, eset.meth$tissue == t]
  
  # Add GABR classes
  fData(eset.sub)$GABA <- gsub("(GABR\\w{1}).*", "\\1", fData(eset.sub)$symbol)
  fData(eset.sub)$GABA <- factor(fData(eset.sub)$GABA)

  # Only include samples with no missing values
  samples <- apply(exprs(eset.sub)[gaba.probes,], 2, function(x) sum(is.na(x)))
  samples <- sampleNames(eset.sub)[which(samples == 0)]

  gaba.hm <- ggaffy_clustmat(eset.sub, what = "rows",
    samples = samples, probesets = gaba.probes, color = "GABA")

    # Swap probesets for symbols
    gaba.hm <- gaba.hm + 
      scale_x_discrete(breaks = gaba.df$id, labels = gaba.df$symbol) +
      scale_y_discrete(breaks = gaba.df$id, labels = gaba.df$symbol) +
      scale_fill_gradient2(midpoint = 1) +
      opts(title = t, legend.position = "none")
    gaba.hms[[t]] <- gaba.hm
}

ggaffy:::ggaffy_layout(2, 2, gaba.hms)
```
