# Tissue-based methylation differences

Identify loci with significantly different levels of methylation between tissue samples collected from each of the 4 brain brain regions.

* results: [`tissue-based-differential-methylation.csv`](https://github.com/aaronwolen/gibbs-brain-project/blob/master/results/tissue-based-differential-methylation.csv)

```{r setup, include=FALSE}
# source("functions/knit_wiki.r")
# knit_wiki("tissue-based-differential-meth-report.Rmd", output.filename = "Cross-tissue methylation differences")

report.title <- "tissue-based-meth-differences"
fig.dir <- file.path("figures", report.title)
fig.prefix <- "meth-tissue-"

dir.create(fig.dir, showWarnings = FALSE)

opts_knit$set(upload.fun = function(file, 
  key = "ec071b3a0c69857c0db26a477c1a568c") imgur_upload(file, key))

opts_chunk$set(fig.path = file.path(fig.dir, fig.prefix),
  cache.path = paste(".cache-", report.title, "/", sep = ""),
  echo=FALSE, results = 'hide',
  cache=TRUE, fig.width=5.5, fig.height=4, dpi=150)
```

```{r load-libraries, message=FALSE, cache=FALSE}
library(genefilter)

library(plyr)
library(reshape2)
library(ggplot2)

# ggpairs function
devtools::source_gist("3298388")

source("functions/heat_matrix.r")
source("functions/print_xtable.r")
source("functions/save-figures.r")
source("functions/plot_probe.r")
```

```{r load-data, cache=FALSE}
source("4b-load-meth-data.r")
```

 
## Methods

### Non-specific filtering

```{r filter-probes}
# Calculate SD for each probe
fData(eset.meth)$sd <- rowSds(exprs(eset.meth), na.rm = TRUE)

# Define probe inclusion threshold
sd.thresh <- 0.1

par(mar = c(4.1, 4.1, 1, 1), xaxs = "i", yaxs = "i")
ax.lim <- c(0, max(fData(eset.meth)$sd * 1.05))
layout(matrix(1:2, ncol = 1), heights = c(.6, .4))

hist(fData(eset.meth)$sd, xlab = "", main = "", xlim = ax.lim)
abline(v = sd.thresh, col = "red")

boxplot(fData(eset.meth)$sd, horizontal = TRUE, xlab = "Probe SD", ylim = ax.lim)
abline(v = sd.thresh, col = "red")

# Vector of probes to include in differential analysis
test.probes <- featureNames(eset.meth)[fData(eset.meth)$sd > sd.thresh]
```

Low-variance probes were filtered out prior to conducting the differential expression analysis. Only probes with an experiment-wide SD > `r sd.thresh` were included. Of the `r nrow(eset.meth)` probes, `r length(test.probes)` exceeded this threshold and were included in the analysis.

### t-test
The t.test was carried out using the `rowttests` from the `genefilter` package. Multiple testing corrections were performed within each region using the Benjamini & Hochberg method for generating q-values.

```{r ttest, dependson="filter-probes"}
tissues <- levels(eset.meth$tissue)

# All combinations of tissue pairs
pairs <- t(combn(tissues, 2))
pair.labels <- apply(pairs, 1, paste, collapse = "-")

t.results <- list()

# Loop through each tissue pair and perform ttest
for(i in seq_len(nrow(pairs))) {
  pair.label <- paste(pairs[i, ], collapse = "-")
  
  # Subset for current tissue pair
  meth.sub <- eset.meth[test.probes, eset.meth$tissue %in% pairs[i,]]
  meth.sub$tissue <- factor(meth.sub$tissue)
  
  t.results[[pair.label]] <- rowttests(meth.sub, "tissue")
  t.results[[pair.label]] <- data.frame(pair = pair.label, 
    id = rownames(t.results[[i]]), t.results[[i]])
}

t.results <- data.frame(do.call("rbind", t.results), row.names = NULL)


# Perform t-test individiually on probes with missing values
na.ts <- which(is.na(t.results$statistic))

pb <- txtProgressBar(max = length(na.ts), style = 3)
for(i in na.ts) {
  setTxtProgressBar(pb, value = which(na.ts == i))
  tissues <- strsplit(as.character(t.results$pair[i]), split = "-")[[1]]
  id <- as.character(t.results$id[i])

  eset.sub <- eset.meth[, eset.meth$tissue %in% tissues]
  
  t.data <- data.frame(exp = exprs(eset.sub)[id, ], tissue = eset.sub$tissue)
  # Remove NA values
  t.data <- t.data[!(is.na(t.data$exp) | is.na(t.data$tissue)),]
  
  t.out <- rowttests(matrix(t.data$exp, nrow = 1), fac = factor(t.data$tissue))
  t.results[i, c("statistic", "dm", "p.value")] <- t.out
  if(any(is.na(t.results[i,]))) {
    stop(paste(i, "produced NAs"))
  }
}


# Correct for multiple testing
t.results$q.value <- unlist(
  with(t.results, tapply(p.value, pair, p.adjust, method = "BH")))


# Export t-test results
write.csv(arrange(t.results, q.value), 
  "results/tissue-based-differential-methylation.csv", row.names = FALSE)

# Annotate t-test results
f.cols <- c("id", "symbol", "chr", "position", "strand", "description")
t.results <- merge(fData(eset.meth)[, f.cols], t.results, by = "id")
t.results <- arrange(t.results, q.value)
```

## Results

```{r results-table, results='asis', dependson="ttest"}
results.table <- ddply(t.results, .(pair), summarise, 
  'p-value < 0.05' = sum(p.value < 0.05, na.rm = TRUE),
  'q-value < 0.05' = sum(q.value < 0.05, na.rm = TRUE))

print_xtable(results.table)
```

### Volcano plot
Volcano plot of mean cross-tissue methylation differences plotted against raw t-test p-values. As expected, the most robust differences are observed in any of the tissue pairs that include the cerebellum, while the fewest differences exist between the two cortical regions. 

```{r volcano-plot, dependson="ttest"}
ggplot(t.results) +
  aes(x = dm, y = -log10(p.value)) +
  geom_point(size = 1) +
  facet_wrap(~pair, nrow = 2) +
  xlab("Mean cross-tissue difference") + 
  ylab(expression(-log[10] * " p-value"))
```


### Pairwise comparisons of mean tissue differences
Pairwise comparison of tissue-based methylation differences across all tissue pairs. In all three tissue combinations that include cerebellum, the methylation differences are highly correlated. The comparisons including pons and either of the two cortical regions are also correlated (the correlation direction is a result of the order in which tissues were compared). 

```{r dm-tissue-pairs, fig.height=5.5, dependson="ttest"}
t.dm <- dcast(t.results, id ~ pair, value.var = "dm")

ggpairs(data = t.dm, 
  facet.vars = pair.labels, alpha = 0.5, size = 0.75) + 
  stat_smooth(method = "lm") + 
  xlab("Mean tissue difference")  + ylab("Mean tissue difference") +
  coord_fixed()
```

#### Mean difference correlations across tissues
The following heatmap depicts the cross-tissue correlations.

```{r dm-tissue-pairs-correlations, fig.width=3.5, fig.height=2.5, dependson="ttest"}
dm.cors <- abs(cor(t.dm[, pair.labels]))
dimnames(dm.cors) <- lapply(dimnames(dm.cors), function(x) sub("-", "\n", x))

heat_matrix(dm.cors, symmetric = TRUE, legend.title = "cor", font.size = 3)
```

### Dot plots for top probes
Dot plots for the 10 most significant probes for each tissue pair are provided [here](https://github.com/aaronwolen/gibbs-brain-project/blob/master/figures/tissue-based-meth-differences/meth-tissue-top-probes.pdf).

```{r sig-probes-dotplots, include=FALSE, dependson="ttest"}
# Identify ten most significant probes for each tissue pair
top.probes <- melt(sapply(pair.labels, simplify = FALSE, 
  FUN = function(x) t.results$id[t.results$pair == x][1:10]))
names(top.probes) <- c("id", "pair")
  
probe.plots <- list()
  
for(i in 1:nrow(top.probes)) {
  cur.probe <- as.character(top.probes$id[i])
  cur.pair <- as.character(top.probes$pair[i])
  
  # Index of corresponding row in t.results 
  r.id <- which(t.results$id == cur.probe & t.results$pair == cur.pair)
  
  title <- paste(cur.probe, " (", cur.pair, " q-value = ",
    signif(t.results$q.value[r.id], 2), ")", sep = "")
  
  probe.plots[[cur.probe]] <- probe_dotplot(eset.meth, probe = cur.probe, 
    x = "tissue", point.label = "individual", color = "gender") +
  stat_summary(fun.data = function(x) mean_sdl(x, mult = 1), 
      geom = "pointrange", color = "blue", alpha = 0.5) +
  opts(title = title, axis.title.x = theme_blank(), legend.position = "right") +
  guides(color = guide_legend(override.aes = list(size = 4, alpha = 1)))
}
  
pdf(file.path(fig.dir, paste(fig.prefix, "top-probes.pdf", sep = "")),
  width = 7, height = 6)
temp <- lapply(probe.plots, print)
dev.off()
```  
  